\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2024-05-28 doc/paper-jss/sdmTMB-paper.Rnw
Revisions

We can test for spatial autocorrelation with a visual inspection or a statistical test of the residuals.
Here, we demonstrate an approach using an implementation of Moran's I from the \pkg{ape} package \citep{gittleman1990, ape}.
\Rev{A3}We set \code{type = "mle-mvn"} to denote setting fixed effects at their maximum likelihood estimates but taking a single draw from the approximate (multivariate normal) distribution of the random effects \citep{waagepetersen2006, thygesen2017a}.

@@

We can also see that the marginal AIC of the model with spatial random fields is
lower:

@@

@@

We can do additional model comparison with cross validation via \fct{sdmTMB\_cv}.
We will do 10-fold cross validation with the folds constructed randomly. We will set the seed each time to ensure the folds are consistent.
Using the \code{fold\_ids} argument, we could supply our own folds and conduct spatially blocked cross validation. If we set a parallel plan with the \pkg{future} package \citep{bengtsson2021future}, our folds will be fit in parallel.

@@

Indeed, the log likelihood of left-out data is considerably higher for the model that includes random fields indicating better out-of-sample predictions:

@@

In practice, we would repeat this procedure several times to ensure the rank order is not sensitive to the randomly chosen folds and, if it was, consider averaging across multiple folds or increasing the number of folds.

@@

% TODO: data citation!?
\sloppy To visualize our model, we can make predictions with the \fct{predict} method (\code{?predict.sdmTMB}) and optionally use the \code{newdata} argument to predict on a new data frame containing locations and all predictor columns.
Here, we will predict on a 2 \(\times\) 2 km grid (\code{qcs\_grid}) that covers the entire region of interest so we can visualize the predictions spatially.
The grid contains spatial covariate columns and all predictors used in the model set at values for which we want to predict.  Some covariates might be fixed at a specified value for all predictions, such that we are predicting the expected value for samples conditional on those specified values.
We call these covariates \emph{detectability} or \emph{catchability} covariates, given the model is predicting the target variable while controlling for the additional influence of these detectability covariates that was estimated in fitted data \cite{thorson2019}.
The output of \fct{predict} is a data frame containing overall estimates in link space (\code{est}), estimates from the non-random-field components (\code{est\_non\_rf}; here, intercept and depth), and estimates from the individual random field components (here, \code{omega\_s}---the spatial field).
% The overall predictions combining fixed and random effects are plug-in predictions.
We can plot these with \fct{geom\_raster} from the \pkg{ggplot2} package (Figure~\ref{fig:pcod-predict-maps}).

@@

<<dog-utms, echo=TRUE>>=
dat <- add_utm_columns(dat, c("lon", "lat"),
  units = "km", utm_crs = 32609)
dat$log_depth <- log(dat$depth)
mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 8)
@

@@

Our model can be written as
\[
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
\exp \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
O_{\bm{s},t} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \gamma_t +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t}
\right),\\
\bm{\omega} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_\omega \right),\\
\bm{\epsilon}_{t} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_{\epsilon} \right),
\end{aligned}
\]
where \(\bm{\beta}\) are coefficients associated with the main effects, \(O_{\bm{s},t}\) represents the offset (here, log area swept), \(\gamma_t\) represents the time-varying coefficients, \(\omega_{\bm{s}}\) is the value from a spatial field (representing constant latent spatial effects) and \(\epsilon_{\bm{s},t}\) is a value from a spatiotemporal field  (representing latent spatial effects that vary by year).
The temporally varying intercepts \(\gamma_t\) are modeled as a stationary AR(1) process,
\[
\begin{aligned}
  \gamma_{t=1} &\sim \operatorname{Normal} \left(0, \sigma^2_{\gamma} \right),\\
  \gamma_{t>1} &\sim \operatorname{Normal} \left(\rho_\gamma\gamma_{t-1}, \sqrt{1 - \rho_\gamma^2} \sigma^2_{\gamma} \right),
\end{aligned}
\]
where \(\rho_\gamma\) represents the correlation between intercepts \(\gamma\) at time $t-1$ and $t$.

@@

<<dog-tw, results='hide', message=FALSE, warning=FALSE, echo=TRUE>>=
fit_tw <- sdmTMB(
  catch_weight ~ s(log_depth),
  data = dat,
  mesh = mesh,
  family = tweedie(),
  offset = log(dat$area_swept),
  time = "year",
  time_varying = ~ 1,
  time_varying_type = "ar1",
  spatial = "on",
  spatiotemporal = "iid",
  anisotropy = TRUE,
  silent = FALSE
)
@

@@

<<dog-update, results='hide', message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE>>=
fit_dg <- update(fit_tw, family = delta_gamma())
fit_dl <- update(fit_tw, family = delta_lognormal(),
  control = sdmTMBcontrol(newton_loops = 2))
fit_dpg <- update(fit_tw, family = delta_gamma(type = "poisson-link"))
fit_dpl <- update(fit_tw, family = delta_lognormal(type = "poisson-link"))
@

>>>>

We can test for spatial autocorrelation with a visual inspection or a statistical test of the residuals.
Here, we demonstrate an approach using an implementation of Moran's I from the \pkg{ape} package \citep{gittleman1990, ape}.
\Rev{A3}We set \code{type = "mle-mvn"} to denote setting fixed effects at their maximum likelihood estimate (MLE) but taking a single draw from the approximate (multivariate normal) distribution of the random effects \citep{waagepetersen2006, thygesen2017a}.

@@

We can also see that the marginal Akaike information criterion (AIC) \citep{akaike1974} of the model with spatial random fields is lower:

@@

\Rev{A1}We note that caution is warranted in performing model selection via marginal AIC for models involving penalized smoothing (spatial or otherwise) \citep[e.g.][]{greven2010, safken2021}.
In marginal AIC calculation, the degrees of freedom is based on the number of fixed effects and does not account for the degree of random effect penalization as conditional AIC would.
Methods to estimate effective degrees of freedom were recently demonstrated in \citep{thorsonXX}, but are not yet included in \sdmTMB.

@@

As an alternative to AIC, we can conduct model comparison with cross validation via \fct{sdmTMB\_cv}.
We will do 10-fold cross validation with the folds constructed randomly.
We will set the seed each time to ensure the folds are consistent.
Using the \code{fold\_ids} argument, we could supply our own folds and conduct spatially blocked cross validation \citep[][]{roberts2017}.
If we set a parallel plan with the \pkg{future} package \citep{bengtsson2021future}, our folds will be fit in parallel.

@@

We can then calculate any performance metric of interest for comparison.
\Rev{A2}A common metric would be the log score or log predictive (likelihood) density (lpd) of the left-out data \citep[][]{geisser1979, vehtari2017}

$$
\mathrm{lpd} = \sum_{i=1}^{n} \ell (\widehat{y_i} | y_i),
$$

where $n$ represents the number of left-out data points, $\ell$ represents the log likelihood, $y_i$ represents left-out data point $i$, and $\widehat{y_i}$ represents the prediction for left-out data point $i$ with the fixed effects at their MLE and random effects at their Empirical Bayes estimate.

Indeed, the log likelihood predictive density for the left-out data is considerably higher for the model that includes random fields indicating better out-of-sample predictions:

@@

In practice, we would repeat this procedure several times to ensure the rank order is insensitive to the randomly chosen folds and, if it was, consider averaging across multiple folds or increasing the number of folds.

@@

\sloppy To visualize our model, we can make predictions with the \fct{predict} method (\code{?predict.sdmTMB}) and optionally use the \code{newdata} argument to predict on a new data frame containing locations and all predictor columns.
Here, we will predict on a 2 \(\times\) 2 km grid (\code{qcs\_grid}) that covers the entire region of interest so we can visualize the predictions spatially.
The grid contains spatial covariate columns and all predictors used in the model set at values for which we want to predict.  Some covariates might be fixed at a specified value for all predictions, such that we are predicting the expected value for samples conditional on those specified values.
We call these covariates \emph{detectability} or \emph{catchability} covariates, given the model is predicting the target variable while controlling for the additional influence of these detectability covariates that was estimated in fitted data \cite{thorson2019}.
The output of \fct{predict} is a data frame containing overall estimates in link space (\code{est}), estimates from the non-random-field components (\code{est\_non\_rf}; here, intercept and depth), and estimates from the individual random field components (here, \code{omega\_s}---the spatial field).
% The overall predictions combining fixed and random effects are plug-in predictions.
We can plot these with \fct{geom\_raster} from the \pkg{ggplot2} package (Figure~\ref{fig:pcod-predict-maps}).

@@

<<dog-utms, echo=TRUE>>=
dat <- add_utm_columns(dat, c("lon", "lat"),
  units = "km", utm_crs = 32609)
dat$log_depth <- log(dat$depth)
mesh <- make_mesh(dat, xy_cols = c("X", "Y"), n_knots = 200)
@

@@

Our model can be written as
\[
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
\exp \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
O_{\bm{s},t} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \gamma_t +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t}
\right),\\
\bm{\omega} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_\omega \right),\\
\bm{\epsilon}_{t} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_{\epsilon} \right),
\end{aligned}
\]
where \(\bm{\beta}\) are coefficients associated with the main effects, \(O_{\bm{s},t}\) represents the offset (here, log area swept), \(\gamma_t\) represents the time-varying coefficients, \(\omega_{\bm{s}}\) is the value from a spatial field (representing constant latent spatial effects) and \(\epsilon_{\bm{s},t}\) is a value from a spatiotemporal field  (representing latent spatial effects that vary by year).
The temporally varying intercepts \(\gamma_t\) are modeled as a stationary AR(1) process,
\[
\begin{aligned}
  \gamma_{t=1} &\sim \operatorname{Normal} \left(0, \sigma^2_{\gamma} \right),\\
  \gamma_{t>1} &\sim \operatorname{Normal} \left(\rho_\gamma\gamma_{t-1}, \sqrt{1 - \rho_\gamma^2} \sigma^2_{\gamma} \right),
\end{aligned}
\]
where \(\rho_\gamma\) represents the correlation between intercepts \(\gamma\) at time $t-1$ and $t$.
We include the argument \code{extra\_time}, which represents all time steps to include in the latent process, such that any autoregressive processes represent annual time steps as opposed to the irregular years of data collection.

@@

<<dog-tw, results='hide', message=FALSE, warning=FALSE, echo=TRUE>>=
fit_tw <- sdmTMB(
  catch_weight ~ s(log_depth),
  data = dat,
  mesh = mesh,
  family = tweedie(),
  offset = log(dat$area_swept),
  time = "year",
  time_varying = ~ 1,
  time_varying_type = "ar1",
  spatial = "on",
  spatiotemporal = "iid",
  anisotropy = TRUE,
  extra_time = seq(min(dat$year), max(dat$year)),
  silent = FALSE
)
@

@@

<<dog-update, results='hide', message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE>>=
fit_dg <- update(fit_tw, family = delta_gamma())
fit_dl <- update(fit_tw, family = delta_lognormal())
fit_dpg <- update(fit_tw, family = delta_gamma(type = "poisson-link"))
fit_dpl <- update(fit_tw, family = delta_lognormal(type = "poisson-link"))
@
